Do researchers preferentially collaborate with same-gendered colleagues?

### 2018-05-27 Gender homophilly project.Rmd
Some writing

---
title: "Do researchers preferentially collaborate with same-gendered colleagues?"
subtitle: Online Supplementary Material
author: "Luke Holman and Claire Morandin"
output:
  html_document: 
    toc: true # show a table of contents
    toc_float: true # make table of contents float at the side of the page
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  # set to true if you want to number sections at each table header
    theme: lumen # Luke's favourite fonts and colours
    code_folding: hide # Add some buttons to show/hide the code
---

@@

```{r warning=FALSE, message=FALSE, results='hide'}
library(dplyr) # 'data wrangling' 
library(stringr) # string manipulation
library(ggplot2) # plots
library(parallel) # parallel computing
library(lme4) # mixed models
library(lmerTest) # mixed models
library(pander) # nice Markdown tables
library(kableExtra) # scrollable table
library(tidyr) # spread and gather functions
library(gridExtra) # arranging ggplots
library(mgcv) # generalised additive models
library(ggbeeswarm) # beeswarm plots
library(Cairo) # for exporting the alpha characters via ggsave

@@

### Function to retrieve data from the SQLite database
Importantly, this function discards all papers where we do not know the gender of every single author. This is conservative but greatly simplifies the test -- we'd have to be careful how we included papers where some author's genders were unknown. Our failures to identify author gender are almost certainly randomly distributed across papers with respect to gender homophily, this omission makes our tests less sensitive but should not bias their outcomes.

@@

  if(over.write | !(file.name %in% list.files())){ # Only do the test if over.write = TRUE or there is no datafile in the working directory

@@

    # Loop over all the journals, try to get the data, and if there is enough, store the data we need as a list entry in 'count.data'
    for(i in 1:length(journals)){

      # Get the relevant data for the focal journal out of the database using dplyr
      # the gender.95 column contains author genders in a format like "MFU" - 
      # MFU means the first author was male, second female, and third was of unknown gender. 
      # Genders were only recorded as M or F if the person's given name is associated with one gender >95% of the time.
      focal <- pubmed_sqlite %>% 
        filter(journal == journals[i]) %>% # restrict to focal journal
        select(gender.95, date) %>%        # get these columns
        collect(n = Inf) %>%               # get the data into memory
        filter(nchar(gender.95) > 1)       # only keep papers with >1 author

@@

      # Restrict the analysis to a specified period of time. My logic is that if the gender ratio changes over time, it will give the false impression of homophily, as explained in Bergstrom et al.'s comment here: http://www.michaeleisen.org/blog/?p=1931
      # For example for papers from the last one year in the dataset, set start.of.time.period = 1 and end.of.time.period = 0
      # For papers from between 5 and 6 yeras ago, set start.of.time.period = 6 and end.of.time.period = 5
      focal <- focal %>% 
        filter(date > (start.of.time.period * -1), 
               date < (end.of.time.period * -1)) %>% 
        select(-date)

@@

          # Make a summary of the number of M and F authors on each paper, and the gender of the first and last author
          n.char <- nchar(focal)
          counts <- data.frame(nMales = str_count(focal, "M"), 
                               nFemales = str_count(focal, "F"),
                               firstF = substr(focal, 1, 1) == "F",
                               lastF = substr(focal, n.char, n.char) == "F")

          rm(focal) # Explicitly discard 'focal', in case this helps to reduce memory footprint

@@

There is a good reason to test whether $\alpha$ is significantly greater (or lower) than expected under the null, rather than simply checking whether the 95% CIs on $\alpha$ overlap zero. The reason is that the null expectation for $\alpha$ is not centred zero for small datasets. This issue was summed up clearly by Prof. Carl Bergstrom in a comment on a blog post (http://www.michaeleisen.org/blog/?p=1931). Prof. Bergstrom wrote:

@@

### Add additional metadata to the $\alpha$ results
Here, we assign extra information to each journal - research discipline (see Holman _et al._ in prep), ISI impact factor (where available), and the country that publishes the journal (where known; information taken from PubMed).
```{r message=FALSE, warning=FALSE}
# Calculate the impact factor of each journal relative to other journals in its discipline (using residuals from a mixed model)
journal.data$logIF <- log10(journal.data$IF) # Take the log10 of the impact factor
journal.data$standardised.IF <- NA
# Calculate residual log10 IF for each journal (i.e. IF relative to the field that it is in)
journal.data$standardised.IF[!is.na(journal.data$logIF)] <- 
  resid(lmer(logIF ~ (1|discipline), data = journal.data[!is.na(journal.data$logIF), ]))

@@

#### Figure showing the average change in $\alpha$ across journals
```{r}
# For each journal where we have observations of alpha' at
# both time points, calculate the change in alpha'
old.vs.new <- master.dataset %>% 
  filter(Subset == "All authors") %>%
  group_by(journal) %>%
  summarise(alpha.old = alpha.adjusted[2], 
            alpha.new = alpha.adjusted[1],
            SE.old = (upper.95.CI2[2] - lower.95.CI2[2]) / 2 * 1.96,
            SE.new = (upper.95.CI2[1] - lower.95.CI2[1]) / 2 * 1.96,
            lower883.old = alpha.old - SE.old * 1.190,
            upper883.old = alpha.old + SE.old * 1.190,
            lower883.new = alpha.new - SE.new * 1.190,
            upper883.new = alpha.new + SE.new * 1.190,
            Significant = "No",
            Significant = replace(Significant, lower883.new > upper883.old, "Yes"),
            Significant = replace(Significant, lower883.old > upper883.new, "Yes")) %>%
  filter(!is.na(alpha.old))

old.vs.new %>%
  ggplot(aes(x = alpha.new - alpha.old, fill = Significant)) + 
  geom_vline(xintercept = 0, linetype = 2) +
  geom_histogram(binwidth = 0.01, colour = "black", fill = "skyblue") + 
  scale_fill_manual(values = c("skyblue", "white")) +
  xlab("Change in \u03B1' over the last ten years") + 
  ylab("Number of journals") +
  theme_minimal()
```
<br></br>
**Figure S`r supp.figure.counter`**: Histogram showing the distribution of differences in $\alpha'$ between the 2015-16 and 2005-6 samples, where positive numbers indicate an increase in $\alpha'$ with time. The mean is slightly greater than 0, indicating a mild increase in average $\alpha'$ with time.

@@

#### Model of the average change in $\alpha$
Journals had slightly lower values of $\alpha'$ ten years ago than they do today. The difference in mean $\alpha'$ values has an effect size of Cohen's d = 0.09, which is considered small. The effect is statistically significant, since we have a large sample of journals. The model includes 'journal' and 'discipline' as random effects, preventing pseudoreplication caused by measuring the same journals twice (i.e. once in each time period), and allowing us to test whether the discipline of the journal explains any variance in $\alpha'$. Both random factors explain only around 1% of the variance. 
```{r}
summary(lmer(scale(alpha.adjusted) ~ Age + (1|journal) + (1|discipline), # Note that the response variable gets scaled, to make the effect size interpretable as Cohen's d
             weights = 1 / ((upper.95.CI2 - lower.95.CI2)/2*1.96), # Weight each observation of alpha by its precision
             data = master.dataset %>% 
               filter(Subset == "All authors", journal %in% old.vs.new$journal)))
```

@@

summary(lmer(scale(alpha.adjusted) ~ Subset + (1|journal) + (1|discipline), 
             weights = 1 / ((upper.95.CI2 - lower.95.CI2)/2*1.96),
             data = first.vs.last.data))
```

@@

### Gender homophily covaries with author gender ratio
Journals with a male-biased author gender ratio showed less author gender homophily, while journals with an even or female-biased author gender ratio showed more homophily. The pattern was similar whether we calculated $\alpha$ for all authors, first authors only, or last authors only. 
```{r fig.width=6, fig.height=5}
gender.ratio <- master.dataset %>% 
  filter(Age == "New") %>% 
  ggplot(aes(x = (gender.ratio.of.sample - 50), y = alpha.adjusted)) + 
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_point(alpha = 0.4, size = 0.7) + 
  stat_smooth(method = "gam", formula = y ~ s(x, k = 3), colour = "tomato", fill = "skyblue") + 
  xlab("Deviation from gender parity for all authors in this journal") + 
  ylab("Coefficient of homophily (\u03B1')") +
  facet_wrap(~Subset) + 
  theme_minimal() +
  theme(legend.position = "none") 
ggsave(plot = gender.ratio, filename = "figures/figure3.pdf", device = cairo_pdf, width=7, height=5)
gender.ratio
```
<br></br>
**Figure `r figure.counter`**: There is a weakly positive, non-linear relationship between the gender ratio of authors publishing in a journal, and the coefficient of homophily ($\alpha'$). Specifically, journals with 50% women authors or higher tended to have more same-sex coauthorships than did journals with predominantly men authors.  This relationship held whether $\alpha'$ was calculated for all authors, first authors only, or last authors only. A negative value on the x axis denotes an excess of men authors, a positive value denotes an excess of women authors, and zero denotes gender parity. The lines were fitted using generalised additive models with the smoothing parameter $k$ set to 3.

>>>>

---
title: "Do researchers preferentially collaborate with same-gendered colleagues?"
subtitle: Online Supplementary Material
author: "Luke Holman and Claire Morandin"
output:
  html_document: 
    toc: true # show a table of contents
    toc_float: true # make table of contents float at the side of the page
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  # set to true if you want to number sections at each table header
    theme: lumen # nice fonts and colours
    code_folding: hide # Add some buttons to show/hide the code
---

@@

```{r warning=FALSE, message=FALSE, results='hide'}
library(dplyr) # data handling 
library(stringr) # string manipulation
library(ggplot2) # plots
library(parallel) # parallel computing
library(lme4) # mixed models
library(lmerTest) # mixed models
library(pander) # nice Markdown tables
library(kableExtra) # scrollable html tables
library(tidyr) # spread and gather functions
library(gridExtra) # arranging ggplots
library(mgcv) # generalised additive models
library(ggbeeswarm) # beeswarm plots
library(Cairo) # for exporting the character 'alpha' via ggsave

@@

### Function to retrieve data from the SQLite database
Importantly, this function discards all papers where we do not know the gender of every single author. This is conservative but greatly simplifies the test -- we'd have to be careful how we incorporated information from papers where some author's genders were unknown. Our failures to identify author gender are almost certainly randomly distributed across papers with respect to gender homophily, and so this omission makes our tests less sensitive but should not bias their outcomes.

@@

  # Only do the test if over.write = TRUE or there is no datafile in the working directory
  if(over.write | !(file.name %in% list.files())){

@@

    # Loop over all the journals, try to get the data, and if 
    # there is enough, store the data we need as a list entry in 'count.data'
    for(i in 1:length(journals)){

      # Get the relevant data for the focal journal out of the database using dplyr
      # the gender.95 column contains author genders in a format like "MFU" - 
      # MFU means the first author was male, second female, and third was of unknown gender. 
      # Genders were only recorded as M or F if the given name is associated with one gender >=95% of the time.
      focal <- pubmed_sqlite %>% 
        filter(journal == journals[i]) %>% # restrict to focal journal
        select(gender.95, date) %>%        # get these columns
        collect(n = Inf) %>%               # get the data into memory
        filter(nchar(gender.95) > 1)       # only keep papers with >1 author

@@

      # Restrict the analysis to a specified period of time. My logic is that if the gender ratio changes over time, it will give the false impression of homophily, as explained in Bergstrom et al.'s comment here: http://www.michaeleisen.org/blog/?p=1931
      # For example for papers from the last one year in the dataset, set start.of.time.period = 1 and end.of.time.period = 0
      # For papers from between 5 and 6 years ago, set start.of.time.period = 6 and end.of.time.period = 5
      focal <- focal %>% 
        filter(date > (start.of.time.period * -1), 
               date < (end.of.time.period * -1)) %>% 
        select(-date)

@@

          # Make a summary of the number of M and F authors on each paper, and the gender of the first and last author
          n.char <- nchar(focal)
          counts <- data.frame(nMales = str_count(focal, "M"), 
                               nFemales = str_count(focal, "F"),
                               firstF = substr(focal, 1, 1) == "F",
                               lastF = substr(focal, n.char, n.char) == "F")
          rm(focal) # Explicitly discard 'focal', to reduce memory footprint

@@

There is a good reason to test whether $\alpha$ is significantly greater (or lower) than expected under the null, rather than simply checking whether the 95% CIs on $\alpha$ overlap zero. The reason is that the null expectation for $\alpha$ is not centred on zero for small datasets. This issue was summed up clearly by Prof. Carl Bergstrom in a comment on a blog post (http://www.michaeleisen.org/blog/?p=1931). Prof. Bergstrom wrote:

@@

### Add additional metadata to the $\alpha$ results
Here, we assign extra information to each journal - research discipline (see Holman _et al._ 2018), ISI impact factor (where available), and the country that publishes the journal (where known; information taken from PubMed).
```{r message=FALSE, warning=FALSE}
# Calculate the impact factor of each journal relative to other journals in its discipline (using residuals from a mixed model)
journal.data$logIF <- log10(journal.data$IF) # Take the log10 of the impact factor
journal.data$standardised.IF <- NA
# Calculate residual log10 IF for each journal (i.e. IF relative to the field that it is in)
journal.data$standardised.IF[!is.na(journal.data$logIF)] <- 
  resid(lmer(logIF ~ (1|discipline), data = journal.data[!is.na(journal.data$logIF), ]))

@@

#### Figure showing the average change in $\alpha$ across journals
```{r}
# For each journal where we have observations of alpha' at
# both time points, calculate the change in alpha'
old.vs.new <- master.dataset %>% 
  filter(Subset == "All authors") %>%
  group_by(journal) %>%
  summarise(alpha.old = alpha.adjusted[2], 
            alpha.new = alpha.adjusted[1],
            SE.old = (upper.95.CI2[2] - lower.95.CI2[2]) / (2 * 1.96),
            SE.new = (upper.95.CI2[1] - lower.95.CI2[1]) / (2 * 1.96),
            lower883.old = alpha.old - SE.old * 1.190,
            upper883.old = alpha.old + SE.old * 1.190,
            lower883.new = alpha.new - SE.new * 1.190,
            upper883.new = alpha.new + SE.new * 1.190,
            Significant = "No",
            Significant = replace(Significant, lower883.new > upper883.old, "Yes"),
            Significant = replace(Significant, lower883.old > upper883.new, "Yes")) %>%
  filter(!is.na(alpha.old))

old.vs.new %>%
  ggplot(aes(x = alpha.new - alpha.old, fill = Significant)) + 
  geom_vline(xintercept = 0, linetype = 2) +
  geom_histogram(binwidth = 0.01, colour = "black", fill = "skyblue") + 
  scale_fill_manual(values = c("skyblue", "white")) +
  xlab("Change in \u03B1' over the last ten years") + 
  ylab("Number of journals") +
  theme_minimal()
```
<br></br>
**Figure S`r supp.figure.counter`**: Histogram showing the distribution of differences in $\alpha'$ between the 2015-16 and 2005-6 samples, where positive numbers indicate an increase in $\alpha'$ with time. The mean is slightly positive (namely `r with(old.vs.new, mean(alpha.new - alpha.old))`), indicating a mild increase in average $\alpha'$ with time.

@@

#### Model of the average change in $\alpha$
Journals had slightly lower values of $\alpha'$ ten years ago than they do today. The difference in mean $\alpha'$ values has an effect size of Cohen's d = 0.09, which is considered small. The effect is statistically significant, since we have a large sample of journals. The model includes 'journal' and 'discipline' as random effects, preventing pseudoreplication caused by measuring the same journals twice (i.e. once in each time period), and allowing us to test whether the discipline of the journal explains any variance in $\alpha'$. Both random factors explain only around 1% of the variance. 
```{r}
# Note that the response variable gets scaled, to make the effect size interpretable as Cohen's d
summary(lmer(scale(alpha.adjusted) ~ Age + (1|journal) + (1|discipline), 
             weights = 1 / ((upper.95.CI2 - lower.95.CI2)/(2*1.96)), # Weight each observation of alpha by 1/SE
             data = master.dataset %>% 
               filter(Subset == "All authors", journal %in% old.vs.new$journal)))
```

@@

summary(lmer(scale(alpha.adjusted) ~ Subset + (1|journal) + (1|discipline), 
             weights = 1 / ((upper.95.CI2 - lower.95.CI2)/(2*1.96)),
             data = first.vs.last.data))
```

@@

### Gender homophily covaries with author gender ratio
Journals with a male-biased author gender ratio showed less author gender homophily, while journals with an even or female-biased author gender ratio showed more homophily. The pattern was similar whether we calculated $\alpha$ for all authors, first authors only, or last authors only. 
```{r fig.width=6, fig.height=5}
gender.ratio <- master.dataset %>% 
  filter(Age == "New") %>% 
  ggplot(aes(x = (gender.ratio.of.sample - 50), y = alpha.adjusted)) + 
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_point(alpha = 0.4, size = 0.7) + 
  stat_smooth(method = "gam", 
              formula = y ~ s(x, k = 3), 
              colour = "tomato", fill = "skyblue") + 
  xlab("% women authors (realtive to gender parity)") + 
  ylab("Coefficient of homophily (\u03B1')") +
  facet_wrap(~Subset) + 
  theme_minimal() +
  theme(legend.position = "none") 
ggsave(plot = gender.ratio, filename = "figures/figure3.pdf", 
       device = cairo_pdf, width=7, height=5)
gender.ratio
```
<br></br>
**Figure `r figure.counter`**: There is a weakly positive, non-linear relationship between the gender ratio of authors publishing in a journal, and the coefficient of homophily ($\alpha'$). Specifically, journals with 50% women authors or higher tended to have more same-sex coauthorships than did journals with predominantly men authors.  This relationship held whether $\alpha'$ was calculated for all authors, first authors only, or last authors only. A negative value on the x axis denotes an excess of men authors, a positive value denotes an excess of women authors, and zero denotes gender parity. The lines were fitted using generalised additive models with the smoothing parameter $k$ set to 3.
